---
title: "Introduction to R Notebooks"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    df_print: paged
params:
  data_path: /share/dsa435601s26/PA-Class/project_data_sets/sample/
---

# What is an R Notebook?

An R Notebook (or R Markdown document) lets you combine **text**, **code**, and **output** all in one document. This is useful because:

- You can explain your thinking alongside your code.
- Your analysis is reproducible -- anyone can re-run it.
- You can produce polished HTML, PDF, or Word reports from a single file.

A `.Rmd` file has three types of content:

1. **YAML header** -- the block at the top between `---` lines that controls the document settings.
2. **Markdown text** -- plain text with simple formatting (like this section).
3. **Code chunks** -- blocks of R code that run and display results.

# Markdown Basics

You can format text using Markdown:

- **Bold** with `**double asterisks**`
- *Italic* with `*single asterisks*`
- `Inline code` with backticks
- Bulleted lists with `-`
- Numbered lists with `1.`, `2.`, etc.
- Headers with `#`, `##`, `###`

# Your First Code Chunk

A code chunk starts with ` ```{r} ` and ends with ` ``` `. When you "knit" the document, R runs each chunk and inserts the output below it.

```{r first-chunk}
# This is a comment -- R ignores lines starting with #
# Let's do some basic arithmetic
2 + 3
10 / 4
```


# Reading Data from a File

In your projects, you will read data from external files.

```{r read-data, eval=FALSE}
# Read a CSV file
my_data <- read.csv(paste0(params$data_path,"sample_students.csv"))

# View the first few rows
head(my_data)

# Check the structure
str(my_data)
```

# Chunk Options

You can control how each chunk behaves by adding options inside `{r}`:

| Option            | Effect                                  |
|-------------------|-----------------------------------------|
| `echo=FALSE`      | Hide the code, show only the output     |
| `eval=FALSE`      | Show the code but don't run it          |
| `message=FALSE`   | Suppress package messages               |
| `warning=FALSE`   | Suppress warnings                       |
| `fig.width=8`     | Set figure width (inches)               |
| `fig.height=5`    | Set figure height (inches)              |

# Knitting Your Document

To produce the final report, click the **Knit** button in RStudio (or press `Ctrl+Shift+K` / `Cmd+Shift+K`). This will:

1. Run all code chunks from top to bottom in a clean environment.
2. Combine the text, code, and output into an HTML file.

**Tip:** If knitting fails, the error message will tell you which chunk caused the problem. Fix that chunk and try again.

# Try It Yourself

```{r practice}
# Try writing your own code here!
# For example, create a vector of your favorite numbers and compute the mean.

```
