---
title: "DSA 495 601 Project Milestone 3"
author: "Your Name"
output:
  html_notebook:
    toc: true
    toc_depth: 1
    number_sections: false
    code_folding: hide
  html_document:
    toc: true
    toc_depth: '1'
    df_print: paged
editor_options:
  chunk_output_type: inline
  markdown: 
    wrap: 72
params:
  data_path: /share/dsa495601f25/project_data_sets/FOLDER_FOR_YOUR_DATA/ # @TODO Select folder within project_data_sets
  my_path: /share/dsa495601f25/YOUR_FOLDER/ # @TODO Select your personal folder
---

For Milestone 3, you will prepare your data for predictive analytics.
This R notebook provides a review of the data preparation concepts we
discuss in class and helps you implement them. Please read through the
entire document closely. You may not need to carry out all steps (e.g.
if you do not have missing values, skip over the step that addresses
this issue), but you can still learn by reading the notes and reviewing
the code.

Your goals are (1) to create *new versions* of variables that need cleaning or processing (e.g. handling missing values, turning text categorical variables into numeric dummies, scaling) and (2) create *new variables* that summarize other variables in new ways (feature engineering). 

To help you identify places where you need to edit the code to match
your dataset, you can look for "@TODO" in comments. Also, please delete
examples discussed in the text and add your own explanations
(documentation or rationale) for decision-making for your own dataset
and project. Finally, please answer the questions (also indicated with
"@TODO").

After you have edited the code as needed for your data and project goals
and run all code chunks successfully, knit your completed file to a
HTML. Download the HTML to your computer and then print or save it to a
PDF file with the following naming convention:
course_section_semester_assignment_unityID.pdf. For example,
DSC495_601_FA25_Milestone3_keporte2.pdf.

# Set-up

#### Load the necessary packages

No edits are needed in the following chunk. Just run it as is.

```{r loadpackages}
.libPaths(c("/share/dsa495601f25/installed_libraries", .libPaths()))
library(dplyr)
library(tidyr)
library(skimr)   
library(janitor)
library(rsample)
library(fastDummies)
```

#### Read in your data

Here you can copy your code from Milestone 2.

```{r readdata}
# Replace with your file name (path specified in yaml header)
# For CSV
data_file <- paste0(params$data_path,"NAMEOFFILE") # @TODO edit to your filename
df <- read.csv(data_file, sep = ";") %>% janitor::clean_names()
  # note the bank marketing data is separated by ; instead of ,

# OR for RDA
# df <- readRDS(data_file)
# View the name of the object loaded
# ls()

# Check that it loaded ok
dim(df)
head(df)
```

### Do any necessary filtering (to align with scoping)

Most of the time, you do **not** need to filter your data. Filtering
should only be done if your project scope requires you to focus on a
**specific subset of the data**.

⚠️ Important: Filtering reduces the size of your dataset. If you filter
too much, you may not have enough data left to train and test your
model.

```{r Filtering}
# Uncomment below and filter for specific criteria only if needed for your scoping
#df_filtered <- df %>% 
#  filter(VAR == "CRITERIA") # @TODO edit here filter for your criteria

# Check size after filtering
#nrow(df)
```


### Summarize your data

This will help you get you starting in identifying what variables need data cleaning and processing. But you may need to add additional summaries or descriptive analyses. 

```{r Summarize}
# Use skimr package to get summaries of all variables
skim(df)
```

# Step 1: Outcome Preparation & Checks

First, in the empty code chunk below, create your *binary* outcome
variable, if it does not already exist. You may reuse code from
**Milestone 2**, but consider modifying it if needed.

> 🔁 Reminder: Your outcome variable should take on only two values —
> `0` or `1`.

Think carefully about how you define the outcome:

-   For example, you might set the outcome to `1` if a student
    **passed** a course, or alternatively, `1` if a student **did not
    pass** — depending on the question you're trying to answer.
-   Your predictive modeling will estimate the likelihood of the outcome
    equaling `1`.

**Question 1: Describe your outcome to be predicted and what each value
(0 and 1) means.**

Implement coding of your outcome in the chunk below. Also, in the same
code chunk, calculate and display the **prevalence** of your outcome
variable, as you did in **Milestone 2**.

⚠️ Important: Note that this will look at the prevalence in your
filtered data. Be sure that the prevalence is not much more skewed than
the non-filtered data! If it does, please revisit.\*\*


```{r OutcomePrep}
# Create binary (1 or 0) version of your outcome
df <- df %>%
  mutate(outcome = as.numeric(OUTCOMEVARIABLE == "FILL IN")) # @TODO Edit for your data

# Check prevalence (percent of rows where the outcome is 1)
mean(df$outcome, na.rm = TRUE)
```

## Step 2: Data Partitioning

In this step, you'll divide your dataset into two parts:

1.  **Training/Validation set**
2.  **Testing set**

Partitioning your data *before* any further preparation is crucial to
prevent **data leakage** — where information from the test set
unintentionally influences model training or feature engineering. To
avoid this, all remaining data preparation steps will be performed
*separately* on the training/validation and testing sets.

### Random Splitting and Stratification

If your data is not organized over time, or if you're looking for a
straightforward approach, you can **randomly split** your dataset.

-   If your outcome variable is **imbalanced** (e.g., very few positive
    cases), consider **stratifying** your split by the outcome class to
    maintain proportional representation in both sets.
-   You may also choose to stratify by other variables if they are
    important and have rare values. Talk to me if you have questions
    about this.

### Splitting by Time

If your data spans across time (e.g., terms in the OULAD dataset or
inspection dates in the restaurant dataset), a **time-based split** may
be more appropriate. In this approach, you save *later observations* for
testing, simulating a real-world scenario where models are trained on
past data and applied to future data. But make sure your test set is
still representative of your training/validation set (not a differnt
time of the year). Reach out to me if you have questions about deciding
on this or if you need help with coding it.

**Question 2: Did you use random splitting or time-based splitting? Why?
Did you stratify your split? If so, by which variable(s), and why?**

### Code Example: Splitting and Saving Your Data

Use the following code to split your dataset:

```{r DataSplit}
# For random splitting
split <- initial_split(df, prop = 0.70, strata = NULL) # @TODO edit tovariable to stratify by if not NULL
trainDat <- training(split)
testDat  <- testing(split)

# For other splitting 
#trainDat <- df %>%
#  filter(code_presentation %in% c("2013B","2013J")) # @TODO this is an example for OULAD, edit for your data
#testDat <- df %>%
#  filter(code_presentation %in% c("2014B","2014J")) # @TODO this is an example for OULAD, edit for your data

# Verify train/test data split
# Check number of rows in training dataset
nrow(trainDat)

# Check number of rows in test dataset  
nrow(testDat)

# Verify that train + test equals original dataset (should be TRUE)
nrow(trainDat) + nrow(testDat) == nrow(df)

# Calculate proportion of data used for training
nrow(trainDat) / nrow(df)

```

# Step 3: Data Processing for Predictive Analytics

After creating your outcome variable and partitioning your data into
training and testing sets, data processing is crucial for model
performance. This section will walk you through the essential steps.

## (a) Handling Missing Values

When building predictive models, it’s important to handle missing values
carefully. Missingness itself can contain useful information — so we
don’t always want to just drop or fill in the missing values without
keeping track of them.

We discussed two simple options for dealing with missing values in
numeric (continuous) variables, and one for categorical variables.

🔹 Option 1: Dummy Variable Adjustment (for numeric variables)

This approach lets you keep the numeric variable, but also capture
whether it was missing.

Steps:

-   Create a new variable that flags whether the value was missing (1 =
    missing, 0 = not missing)

-   Fill in the missing values using the mean of the variable

-   Later, you will include both variables in your model

Example:

If your variable is testscore, you would:

-   Create testscore_missing = 1 if missing, 0 otherwise

-   Create testscore_i = original value if available, mean value if
    missing

-   Use both testscore_i and testscore_missing in your model

```{r ContinuousMissing}
#' This function implements dummy variable adjustment for handling missing data in predictive modeling.
#' For each variable specified, it creates a missing indicator and imputes missing values with the mean.
#' 
#' @param data A data frame containing the variables to process
#' @param var_names A character vector of variable names to process for missing values
#' 
#' @return A data frame with:
#'   - Original variables with missing values imputed with their respective means
#'   - New indicator variables (suffix "_missing") for each processed variable
#'   
#' @details For each variable in var_names, the function:
#'   1. Calculates the mean of non-missing values
#'   2. Creates a missing indicator (1 = missing, 0 = observed)
#'   3. Imputes missing values with the calculated mean
#'   
#' @note This should be applied separately to training and test data

process_missing <- function(data, var_names) {
 for (var in var_names) {
   var_mean <- mean(data[[var]], na.rm = TRUE)
   missing_indicator <- is.na(data[[var]])
   data[[var]][missing_indicator] <- var_mean
   data[[paste0(var, "_missing")]] <- as.numeric(missing_indicator)
 }
 return(data)
}

# Example implementation
vars_to_process <- c("prev_GPA_SS","prev_testscore") # @TODO edit to continuous variables in your data that you want to process in this way, then the remainder of the code should work
trainDat <- process_missing(trainDat, vars_to_process)
testDat <- process_missing(testDat, vars_to_process)

```

🔹 Option 2: Convert Continuous Numeric Variables to Categories

This approach works well when the variable has meaningful thresholds
(e.g., test score cutoffs).

Example:

Instead of using raw scores, you might create:

-   "Below Basic" for scores 0–60

-   "Basic" for scores 61–80

-   "Proficient" for scores 81–100

-   "Missing" for missing values

This captures non-linear relationships and makes missingness just
another category.

This method may lose some detail, but it's a useful tradeoff when you
want simpler variables with meaning.

```{r MakeCategorical}
# Convert numeric variable to meaningful categories with "Missing"
convert_numeric_to_category <- function(vec, breaks, labels) {
  # Create categories using cut
  cat_var <- cut(vec, breaks = breaks, labels = labels, include.lowest = TRUE)
  
  # Add a "Missing" category
  cat_var <- addNA(cat_var)
  levels(cat_var)[is.na(levels(cat_var))] <- "Missing"
  
  return(cat_var)
}

# Example: Binning 'testscore' into performance categories
trainDat$testscore_cat <- convert_numeric_to_category(
  vec = trainDat$testscore, # @TODO edit to match a variable in your data if desired/applicable
  breaks = c(-Inf, 60, 80, Inf), # @TODO edit to define breaks based on the distribution of your data
  labels = c("Below Basic", "Basic", "Proficient") # @TODO edit to match your variable
)
# Repeat exact same in Test Data. Use same breaks and labels as for the training data
testDat$testscore_cat <- convert_numeric_to_category(
  vec = testDat$testscore, # @TODO edit to match a variable in your data if desired/applicable
  breaks = c(-Inf, 60, 80, Inf), # @TODO edit to define breaks based on the distribution of your data
  labels = c("Below Basic", "Basic", "Proficient") # @TODO edit to match your variable
)
# @TODO copy and paste the above code and edit for all variables you want to repeat this process for

```

🔹 Option 3: Add a "Missing" Category (for existing categorical
variables)

For categorical variables, simply add a "Missing" category. That way,
missing values are treated like any other level. However, this can be
handled simultaneously with "one-hot coding" in the next section.
Therefore, if you have a categorical variable with missing values, this
will be addressed below.

**Note**: If you have just a few missing values for a variable, then
your category for missing will be very tiny. You can then combine the
missing category with another category. For example, if you have 3
missing values for gender, then I would recommend, that you have one
category as female and one category that contains both male and missing.

**Question 3: Do you have missing values? If so, which option(s) did you
use to handle missing values (dummy variable adjustment, binning, or
“Missing” category)? Why did you make the choices you did? Do you think
missingness itself might be predictive in your dataset? Explain
briefly.**

## (b) One-Hot Encoding (Dummy Variables) for Categorical Variables

Some machine learning models — like `glmnet` (for lasso, ridge), `keras`
(for neural networks), or support vector machines — **only work with
numeric data**. This means we need to turn categorical variables into
numbers.

The most common way to do this is called **one-hot encoding** (or
**dummy coding**). This creates a new column for each category in your
variable, with values of `1` (yes) or `0` (no).

As noted above, this procedure will also deal with missing values in
categorical variables because it will create a missing value for NA or
other indicator of missing.

### ✅ Dummy Coding with `fastDummies::dummy_cols()`

Let’s say you want to create dummy variables for:

-   `"school_type"`
-   `"testscore_cat"`

Here’s the code:

```{r OneHotCoding}
# Specify columns of categorical variables you want to one hot code
cat_columns = c("marital", "education","contact") # @TODO edit for variables in your data

trainDat <- fastDummies::dummy_cols(
  trainDat,
  select_columns = cat_columns, 
  remove_selected_columns = TRUE
)

testDat <- fastDummies::dummy_cols(
  testDat,
  select_columns = cat_columns,
  remove_selected_columns = TRUE
)
```

This will automatically create new columns like:

-   `school_type_Public`, `school_type_Private`
-   `testscore_cat_Low`, `testscore_cat_Medium`, etc.

Each is a **0/1 column**, and the original variable is removed.

### Handling Missing Values

By default, fastDummies::dummy_cols() treats NA as its own category. It
creates a dummy column called something like var_NA. That column is 1 if
the value was missing, 0 otherwise.

### ⚠️ Be Careful with Numeric Variables

`dummy_cols()` will also create dummy variables for numeric columns —
**but that’s not always what you want.**

> ❌ **Don’t use dummy_cols() on numeric variables like GPA, test score,
> or age** if they have lots of different values. This will create a
> dummy column for every unique number, which is not useful.

**Question 4: Which variables did you one-hot encode? Why?**

## (c) Scaling

Often, variables used in machine learning models have different ranges.
Without scaling, variables with larger absolute values will dominate the
prediction outcome regardless of their actual importance. To give each
feature a fair opportunity to contribute to the prediction outcome, we
must bring all features to a comparable scale.

The two common scaling techniques are:

1.  **Standardization (Z-score Scaling)**: This technique transforms
    features to have zero mean and unit variance by subtracting the mean
    and dividing by the standard deviation. This is generally the
    recommended approach because:
    -   It preserves the shape of the feature's distribution while
        making units comparable
    -   It works well with regularization since all features are
        centered around zero (which we will discuss later)
    -   It handles outliers better than other methods
    -   Most statistical learning methods are developed assuming
        standardized features
2.  **Normalization (Min-Max Scaling)**: This technique scales feature
    values to a fixed range of [0,1]. While less commonly needed,
    normalization is useful in specific cases:
    -   When working with neural networks, particularly in the output
        layer when predicting probabilities
    -   When working with algorithms that use distance measures and you
        want to ensure equal distance weight (e.g., KNN)

**Important Notes:**

-   For most predictive modeling tasks, standardization is the safer
    default choice.

-   Tree-based algorithms (decision trees, random forest, XGBoost) don't
    require scaling, but since we will be using a variety of ML
    algorithms, we will prepare data as such.

-   For highly skewed features, consider log transformation before
    scaling

Example code for standardization:

```{r standardize}
#' Standardize Features for Predictive Modeling
#'
#' @param data A data frame containing features to standardize
#' @param features Character vector of feature names to standardize
#' @param suffix Character suffix for new columns (defaults to "_std")
#' @return Original data frame with added standardized columns
#'
standardize_features <- function(data, features, suffix = "_std") {
 
 # Calculate means and sds for all features at once
 scaling_params <- data %>%
   summarise(across(all_of(features),
                   list(mean = ~mean(., na.rm = TRUE),
                        sd = ~sd(., na.rm = TRUE)))) %>%
   pivot_longer(everything(),
               names_to = c("feature", "stat"),
               names_sep = "_") %>%
   pivot_wider(names_from = stat, values_from = value)
 
 # Add standardized versions
 result_data <- data %>%
   mutate(across(all_of(features),
                ~(. - scaling_params$mean[scaling_params$feature == cur_column()]) /
                  scaling_params$sd[scaling_params$feature == cur_column()],
                .names = "{.col}{suffix}"))
 
 # Store parameters as attribute
 attr(result_data, "scaling_params") <- scaling_params
 
 return(result_data)
}

# Example usage:
features_to_scale <- c("week1") # @TODO Edit variables to be scaled here. 
trainDat <- standardize_features(trainDat, features_to_scale)
testDat <- standardize_features(testDat, features_to_scale)

```

**Question 5: Which variables did you scale? Describe the scaling you
did.**

## (d) Feature engineering

Predictor variables in machine learning are often called "features."
Feature engineering is the process of transforming raw data into
meaningful inputs that enhance the performance of predictive models. It
involves creating, modifying, or selecting features to better capture
the underlying patterns in the data, leveraging both domain expertise
and statistical techniques. While some of the earlier data processing
steps may overlap with feature engineering, the focus here is on
developing new features from raw data—such as combining variables in
ways that extract predictive information or generating key interaction
terms. This steps takes advantage of domain knowledge and data
experience.

Use this section for feature engineering. Keep in mind that if you have
multiple ideas for feature engineering that seem to overlap, you do not
need to select just one. For example, if you have daily student
attendance, you might create features that summarize attendance as the
total number of days present, calculate the percentage of days attended,
or flag periods of consecutive absences. You can create all of these
features—you don't need to pick just one.

Finally, remember to incorporate the practices outlined above for
handling missing values, scaling, and one-hot encoding to ensure your
features are prepared appropriately for modeling, and remember to
feature engineering identically but separately for training/validation
data and for testing data.

```{r}
# Use this chunk for feature engineering. Feel free to add more chunks too. 
```

**Question 6: Did you create any new features from your raw data? If so,
describe at least one and explain why you thought it might help
prediction.**

# Step 4: Review and check your results

Now that you’ve finished preparing your training and testing data, it’s
important to **review your datasets carefully** before moving on to
modeling. This step helps you confirm that your processing was
successful and that you didn’t accidentally introduce problems.

### ✅ Things to check

1.  **Dimensions**

    -   Did the number of rows stay the same after processing?
    -   Did the number of columns change the way you expected (e.g.,
        more columns after dummy coding)?

2.  **Outcome variable**

    -   Is your outcome variable still present in both training and test
        sets?
    -   Does it still take on only two values (`0` and `1`)?

3.  **Missing values**

    -   Are missing values handled the way you intended (e.g.,
        `_missing` columns created, `_NA` dummies, or binned
        categories)?

4.  **Scaling**

    -   Do your standardized variables have mean ≈ 0 and sd ≈ 1 (in the
        training data)?

5.  **Consistency between train and test**

    -   Do the two datasets have the **exact same variables in the same
        order**?
    -   This is essential for modeling.

### 🛠 Example code to help with checks (add more so that you feel confident)

```{r CheckResults}
### 🛠 Example code to help with checks

# Peek at the first rows
head(trainDat)

# Skim again and focus on new variables
skimr::skim(trainDat)

# Alignment check: make sure train and test have identical variables
all.equal(names(trainDat), names(testDat))
```

### ⚠️ Notes for Students

-   If `all.equal()` returns `"TRUE"`, your train and test sets match.
    🎉

-   If not, it will show which variable names differ.

-   Common reasons include:

    -   A category only appeared in one dataset (dummy coding mismatch)
    -   A variable got dropped by mistake in one dataset but not the
        other

If you see mismatches, **fix them before moving on to modeling**.

**Question 7: Do you feel confident your data is ready for at least a
first round of modeling? Why or why not?**

# Step 5: Save your processed training and testing files

```{r SaveFiles}
# Save you training and testing data files using the file path for dataInputPath
saveRDS(trainDat, file = paste0(params$my_path, "trainDat.rds"))
saveRDS(testDat, file = paste0(params$my_path, "testDat.rds"))

```
