---
title: "DSA 495 601 Project Milestone 2"
author: "Your Name"
output:
  html_document:
    toc: true
    toc_depth: '1'
    df_print: paged
  html_notebook:
    toc: true
    toc_depth: 1
    number_sections: false
    code_folding: hide
editor_options:
  chunk_output_type: inline
params:
  data_path: /share/dsa495601f25/project_data_sets/bank_marketing/
---

For Milestone 2, select your dataset for the project and then complete this markdown. Knit your completed file to a HTML. Download the HTML to your computer and then print or save it to a PDF file with the following naming convention:  course_section_semester_assignment_unityID.pdf. For example, DSC495_601_FA25_Milestone2_keporte2.pdf. 

Due: Septebmer 18 before class. 


# Dataset Selection

**Question:** What dataset did you choose? *(Provide the file name or brief description)*

I chose the dataset: `example_dataset.csv`\
This dataset contains provides the opportunity to learn...

# Project Scoping

#### (a) What is the **binary outcome** you plan to predict?

The binary outcome I'm predicting is:\
`outcome_variable` — where 1 = [positive class], 0 = [negative class]

#### (b) Who is your **target population**?

My target population is:\
To filter for this group, I will subset rows where...

#### (c) **When** are you making the prediction?

I’m making predictions at the point when...\
At this time, I have access to data such as...

#### (d) Why would **predictive analytics** be valuable here?

Predictive analytics could help by...\
The results may be used by...

# Read & Explore Your Dataset

#### Load the necessary packages

```{r loadpackages}
.libPaths(c("/share/dsa495601f25/installed_libraries", .libPaths()))
library(dplyr)
library(skimr)   # For easy summaries
library(janitor) # For formatting object names
```

#### Read in your data

```{r readdata}
# Replace with your file name (path specified in yaml header)
# For CSV
data_file <- paste0(params$data_path,"NAMEOFFILE")
df <- read.csv(data_file, sep = ";") %>% janitor::clean_names()
  # note the bank marketing data is separated by ; instead of ,

# OR for RDA
# load(data_file)
# View the name of the object loaded
# ls()
```

#### How many rows and columns?

```{r dimensions}
dim(df)  # Returns rows and columns
```

#### Make the outcome binary (if needed)

```{r binaryoutcome}
# Example: turn "yes"/"no" into 1/0
df <- df %>%
  mutate(outcome_binary = if_else(outcome == "yes", 1, 0))
```

#### Check distribution of the binary outcome

```{r outcomecheck}
# Replace 'outcome_binary' with your actual column name
df %>%
  count(outcome_binary) %>%
  mutate(percent = n / sum(n) * 100)
```

#### Get a summary of your data

```{r summary}
# Overview using skim
skim(df)

# OR: glimpse for a compact preview
glimpse(df)
```

# Questions About the Data

#### What questions do you have about the data at this time?

Some questions I have include:
